---
title: ""
format: html
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup

requiredPackages <- c("tidyverse", "flextable", "rio", "rstan", "loo", "bayesplot")
for (p in 1:length(requiredPackages)){
  if(require(requiredPackages[p], character.only = TRUE)){
    print(paste0(requiredPackages[p], " is loaded correctly"))
  } else {
    print(paste0("installing ", requiredPackages[p], "..."))
    install.packages(paste0(requiredPackages[p]))
    if(require(requiredPackages[p], character.only = TRUE)){
      print(paste0(requiredPackages[p], " is installed and loaded"))
    } else {
      stop(paste0("could not install ", requiredPackages[p]))
    }
  }
}

proj_seed <- 56024
set.seed(proj_seed)

# Calc EBFMI:
check_energy <- function(stanfit){
  sampler_params <- get_sampler_params(stanfit, inc_warmup=FALSE)
  EBFMI <- rep(0, times = length(sampler_params))
  for (n in 1:length(sampler_params)) {
    energies <- sampler_params[n][[1]][,'energy__']
    numer <- sum(diff(energies)**2) / length(energies)
    denom <- var(energies)
    EBFMI[n] <- numer / denom
  }
  return(EBFMI)
}

# Create folders if they don't exist already:
system("mkdir -p data")
system("mkdir -p draws")

```

## 1. Model overview

The following hierarchical mixed-effects Gaussian model was developed within the Bayesian framework of Stan [@gelman2015; @carpenter2017].
Likelihood was drawn from a Gaussian probability distribution function, expressed by the linear predictor ($\mu_{i}$) composed of fixed and random effects: $$
y_i \sim Normal(\mu_i, \sigma) \\
\mu_i = \beta_0 + \beta_1*Q + \alpha_{sample} + \alpha_{position} + \alpha_{habitat} + \alpha_{site}  
$$

The random effects $\alpha_{sample}$ and $\alpha_{site}$ were specified using a random walk model, treating each subsequent downstream site and within-site sample as a random step from the previous without bias [@codling2008].

## 2. Analysis in Stan

Run the setup chunk, then put the file `xs_3D.csv` into the `data/` directory.

### 2.1. Preparing data for Stan

To prepare the data for a hierarchical analysis, I have made the following assumptions:

-   Column `pos`: the entry `downstream_upstream` consists of a mix of upstream and downstream observations, and should be considered its own category.

-   Column `xs_CN`: the value indicates unmeasured distance from the reach center, and the true value depends on whether the point is upstream or downstream.
    I've coerced this into integers such that the furthermost upstream point (`xs8`) is 1, and the most downstream point is 16 (also `xs8`).

```{r}
#| label: data_prep

# import data:
xs_3D <- read.csv(file = "data/xs_3D.csv", header = T)

xs_3D$pos[xs_3D$pos == "downstream_upstream"] <- "mixed"
xs_3D$xs_CN_us <- xs_3D$xs_CN
xs_3D$xs_CN_us[xs_3D$pos == "upstream"] <- paste0("us_", xs_3D$xs_CN[xs_3D$pos == "upstream"])

xs_3D$samp_no <- as.integer(as.factor(xs_3D$xs_CN_us))
xs_3D$samp_no[xs_3D$pos == "upstream"] <- 9 - xs_3D$samp_no[xs_3D$pos == "upstream"]

xs_3D$xs_N <- gsub(" ", "", xs_3D$xs_N)
xs_3D$site <- stringr::str_extract(xs_3D$out_L, "hx[0-9]+")

samp_no <- xs_3D$samp_no

pos_no <- numeric(nrow(xs_3D))
pos_no[xs_3D$pos == "upstream"] <- 1
pos_no[xs_3D$pos == "mixed"] <- 2
pos_no[xs_3D$pos == "downstream"] <- 3

habitat_no <- as.integer(as.factor(xs_3D$xs_N))
site_no <- as.integer(as.factor(xs_3D$site))

y <- xs_3D$wd_R
C <- scale(xs_3D$CA)
I <- scale(xs_3D$ei)

### assemble into list for Stan:

stan_data <- list(y = y,
                  C = as.numeric(C),
                  I = as.numeric(I),
                  samp_no = samp_no,
                  pos_no = pos_no,
                  habitat_no = habitat_no,
                  site_no = site_no,
                  n_obs = length(y),
                  n_samp = max(samp_no),
                  n_pos = max(pos_no),
                  n_habitat = max(habitat_no),
                  n_site = max(site_no)
                  )

```

### 2.2. Sampling model with Stan

I assessed the model stability and efficiency by inspecting the diagnostic outputs of the model.
Parameters were tuned to achieve the following conditions, based on recommendations by @vehtari2021:

-   No incidence of divergent transitions

-   All parameter $\hat{R}$ values within 0.99 \< $\hat{R}$ \< 1.01 (convergence implied at $\hat{R}$ = 1)

-   All Bulk and Tail Effective Sample Sizes (ESS) \> 400 (100 per chain)

-   EBFMI \> 0.3 for all chains (close to 1 is ideal)

```{r}
#| label: stan_all_effects
#| echo: false

ad <- 0.99
mt <- 12
it <- 1000

# --- Run Stan model to get draws parameters
# Call the rstan package
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# time_start <- Sys.time()
stanfit <- stan(file = "flow_norm_hierarchical.stan", # Stan model for Occurrence Model
                data = stan_data,
                control = list(adapt_delta = ad, max_treedepth = mt),
                #control = list(adapt_delta = 0.99, max_treedepth = 14),
                iter = it,
                chains = 4, 
                seed = proj_seed,
                verbose = F) 
  # 255.748 seconds

check_energy(stanfit) #  0.7844560 0.8553087 0.8657838 0.8861246 All above 0.03 - good!
draws <- rstan::extract(stanfit, inc_warmup = FALSE) # Extract the draws parameters
saveRDS(draws, "draws/C_draws.RData")
  
diag_C <- as.data.frame(monitor(stanfit, warmup = 0, print = FALSE))
diag_C_ESS <- diag_C[c("mean", "sd", "2.5%", "97.5%", "n_eff", "Bulk_ESS", "Tail_ESS", "Rhat")]
  
min(diag_C$Bulk_ESS) # 911 - good!
min(diag_C$Tail_ESS) # 1266 - good!
min(na.omit(diag_C$n_eff))    # 1177 - good!
max(diag_C$Rhat)     # 1.007272 - it will do


```

```{r}
#| label: stan_no_random_eff
#| echo: false

ad <- 0.99
mt <- 12
it <- 1000

# --- Run Stan model to get draws parameters
# Call the rstan package
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# time_start <- Sys.time()
stanfit_nr <- stan(file = "flow_norm.stan", # Stan model for Occurrence Model
                data = stan_data,
                control = list(adapt_delta = ad, max_treedepth = mt),
                #control = list(adapt_delta = 0.99, max_treedepth = 14),
                iter = it,
                chains = 4, 
                seed = proj_seed,
                verbose = F) 
  # 255.748 seconds

check_energy(stanfit_nr) #  0.7844560 0.8553087 0.8657838 0.8861246 All above 0.03 - good!
draws <- rstan::extract(stanfit_nr, inc_warmup = FALSE) # Extract the draws parameters
saveRDS(draws, "draws/C_draws_nr.RData")
  
diag_C <- as.data.frame(monitor(stanfit_nr, warmup = 0, print = FALSE))
diag_C_ESS <- diag_C[c("mean", "sd", "2.5%", "97.5%", "n_eff", "Bulk_ESS", "Tail_ESS", "Rhat")]
  
min(diag_C$Bulk_ESS) # 911 - good!
min(diag_C$Tail_ESS) # 1266 - good!
min(na.omit(diag_C$n_eff))    # 1177 - good!
max(diag_C$Rhat)     # 1.007272 - it will do


```

```{r}
#| label: stan_no_fixed_eff
#| echo: false

ad <- 0.99
mt <- 12
it <- 1000

# --- Run Stan model to get draws parameters
# Call the rstan package
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# time_start <- Sys.time()
stanfit_nf <- stan(file = "flow_norm_randomeff.stan", # Stan model for Occurrence Model
                data = stan_data,
                control = list(adapt_delta = ad, max_treedepth = mt),
                #control = list(adapt_delta = 0.99, max_treedepth = 14),
                iter = it,
                chains = 4, 
                seed = proj_seed,
                verbose = F) 
  # 255.748 seconds

check_energy(stanfit_nf) #  0.7844560 0.8553087 0.8657838 0.8861246 All above 0.03 - good!
draws_nf <- rstan::extract(stanfit_nf, inc_warmup = FALSE) # Extract the draws parameters
saveRDS(draws_nf, "draws/C_draws_nf.RData")
  
diag_C <- as.data.frame(monitor(stanfit_nf, warmup = 0, print = FALSE))
diag_C_ESS <- diag_C[c("mean", "sd", "2.5%", "97.5%", "n_eff", "Bulk_ESS", "Tail_ESS", "Rhat")]
  
min(diag_C$Bulk_ESS) # 911 - good!
min(diag_C$Tail_ESS) # 1266 - good!
min(na.omit(diag_C$n_eff))    # 1177 - good!
max(diag_C$Rhat)     # 1.007272 - it will do


```

### 2.3. Model evaluation

I have evaluated the predictive accuracy of the model by comparing the coefficient of determination ($R^2$), and coefficients for intercept and slope; an intercept equal to 0 and regression slope equal to 1 implies the predicted data predicts observed data with the highest accuracy [@piñeiro2008].
For Bayesian models, the mean of the posterior predicted observations was used as the predictand.

```{r}
#| label: fig-P_vs_O
#| fig-height: 6
#| fig-width: 6

# y <- xs_3D$wd_R
# C <- scale(xs_3D$CA)
# I <- scale(xs_3D$ei)

ypred <- apply(draws$y_pred, 2, FUN = mean)
ypred_nr <- apply(draws_nr$y_pred, 2, FUN = mean)
y_obs <- xs_3D$wd_R

lm1 <- lm(y ~ C)
ypred_lm <- predict(lm1)

par(mfrow = c(2, 2))
plot(y = ypred_lm, x = y_obs, title(main = "Linear Regression:\ny ~ C", adj = 0))
plot(y = ypred_nr, x = y_obs, title(main = "Bayesian GLM:\ny ~ C", adj = 0))
plot(y = ypred, x = y_obs, title(main = "Bayesian GLM:\ny ~ C + rand. effects", adj = 0))

# quantile(draws$beta_Q, probs = c(0.025, 0.5, 0.975))

c(mean(draws$intercept), mean(draws$beta_Q))
summary(lm1)

# mean_samp_eff <- round(apply(draws$alpha_samp, 2, FUN = mean), 3)
# samp_eff <- round(apply(draws$alpha_samp, 2, FUN = quantile, probs = 0.025), 3)

summary(lm(y ~ ypred_lm))
summary(lm(y ~ ypred))


```

```{r}
#| label: looic

library(loo)

draws <- readRDS(file = "draws/C_draws.RData")
draws_nr <- readRDS(file = "draws/C_draws_nr.RData")
draws_nf <- readRDS(file = "draws/C_draws_nf.RData")

loo_compare(list(loo(draws$log_lik), loo(draws_nr$log_lik), loo(draws_nf$log_lik)))
# Obviously the random effects model is a better fit!

```
